Reproducible Research - Peer Assessment 1
========
Author: "Vishnu Chevli"
--------
Date: "Thursday, August 07, 2014"
- - -

### Part 0 : Data preparation

Setting global options before we start working on project
```{r, echo=TRUE}
opts_chunk$set(eacho = TRUE, warning = FALSE, fig.width=12)
```


Following code reads data from *acticity.csv* file
It convert reading to data.table for better processing

```{r}
library(data.table)
reading = read.csv('activity.csv', header= TRUE, stringsAsFactors = FALSE)
reading = data.table(reading)
```

Following set of code convert data into required format:
 1. It converts date from character to POSIXct date 
 2. It matches interval with appropriate minute interval 
 3. Once we have date and time we are calculating timestamp 
 4. After 3rd step we are sorting data using timestamp
 5. Finally we are rearranging columns of data

```{r}
reading$date = as.POSIXct(reading$date,format = '%Y-%m-%d')

#Creating dummy data to match interval against time
dummy.frame = data.table(interval = unique(reading$interval), 
                         time = rep(as.ITime('00:00:00',format='%H:%M:%S'),288) + 
                           (0:287)*5*60)
reading = merge(reading,dummy.frame, by = 'interval')
rm(dummy.frame) #Removing as not required now

#Creating, sorting and reording using timestamp
reading$timestamp = reading$date + reading$time
reading = reading[order(timestamp),]
setcolorder(reading, c('date', 'time', 'timestamp', 'steps', 'interval'))
```

- - -

### Part 1: What is mean total number of steps taken per day?

Following code will summarize reading table and calculate following:
 1. Total steps taken per day
 2. Mean of steps taken per day
 3. Median of steps taken per day
```{r}
summaryByDate = reading[,list(TotalSteps = as.numeric(sum(steps,na.rm = TRUE)),
                             Mean = as.numeric(mean(steps, na.rm = TRUE)),
                             Median = as.numeric(median(steps, na.rm = TRUE))), by = date]
```

#### Question 1. Make a histogram of the total number of steps taken each day
Following code will display historian of total steps taken per day
```{r}
library(ggplot2)
library(scales)
ggplot(data = summaryByDate) + 
  geom_histogram(aes(x = Mean), binwidth = 3, color = 'black', fill = 'light green')
```

#### Question 2. Calculate and report the mean and median total number of steps taken per day
Following code will display mean and median of total steps taken per day
```{r}
#Printing summary table
summaryByDate
```

- - - 

### Part 2: What is the average daily activity pattern?

#### Question 1. Make a time series plot of the 5-minute interval against the average number of steps taken, averaged across all days

Following code will find out average activity and display Timeseries graph for daily activity

```{r}
summaryByTime = reading[,list(Mean = as.numeric(mean(steps, na.rm = TRUE))), by = time]

#Finding Member having maximum avergae number of step
maxSteps = summaryByTime[Mean == max(summaryByTime$Mean),]

#Converting to POSIXct Time for ggplot use
summaryByTime$time = as.POSIXct(summaryByTime$time)

ggplot(summaryByTime) + geom_line(aes(x=time,y=Mean), size = 1, color = 'sea green') +
  labs(x = 'Time', y = 'Average Steps', title = 'Average daily activity pattern') +
  scale_x_datetime(labels = date_format("%H:%M"))
```

#### Question 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
At **`r format(maxSteps$time,'%H:%M')`** time maximum number of steps = **`r round(maxSteps$Mean,0)`**

- - -

### Part 3: Imputing missing values
#### Question 1. Calculate and report the total number of missing values in the dataset
Following code will find out number of missing values for all variables and will display it in tabular manner
```{r}
NACounts = data.frame(NA_Count = apply(reading,2,function(x){sum(is.na(x))}))
NACounts
```

#### Question 2. A strategy for filling in all of the missing values in the dataset.
Here for imputation we are assigning mean value for that time period accross all days to that time period.
For example on 1st Oct 2012 at 00:00 there is not entry for step, 
this would be replaced by mean of steps taken at 00:00 min

#### Question 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
Following code will create a new dataset readingAdj with steps adjusted as new columns AdjSteps.
```{r}
#Finding daily summary
summaryByTime = reading[,list(Mean = as.numeric(mean(steps, na.rm = TRUE))), by = time] 

#Adding time based mean column
readingAdj = merge(reading, summaryByTime, by = 'time')

#If steps value missing adjusting to mean of that time
readingAdj$AdjSteps = ifelse(is.na(readingAdj$steps), readingAdj$Mean, readingAdj$steps)

#Removing Mean column
readingAdj$Mean = NULL

#Sorting and reordining of columns
readingAdj = readingAdj[order(timestamp),]
setcolorder(readingAdj, c('date', 'time', 'timestamp', 'steps', 'interval', 'AdjSteps'))
```

#### Question 4. Comparision of adjusted dataset with original.
Following set of code will find out total steps taken in normal case and after adjustments
We are plotting both histograms side-by-side to make difference visible
```{r, warning=FALSE}
#Finding summary of steps after and before adjustment
adjSummaryByDate = readingAdj[,list(TotalSteps = as.numeric(sum(steps,na.rm = TRUE)),
                                    Mean = as.numeric(mean(steps, na.rm = TRUE)),
                                    Median = as.numeric(median(steps, na.rm = TRUE)),
                                    TotalStepsAdj = as.numeric(sum(AdjSteps,na.rm = TRUE)),
                                    MeanAdj = as.numeric(mean(AdjSteps, na.rm = TRUE)),
                                    MedianAdj = as.numeric(median(AdjSteps, na.rm = TRUE))), by = date]

#Preparing a new variable for histogram
forGraph = data.table(TypeData = c(rep(c('Original','Adjusted'),each = nrow(adjSummaryByDate))),
                      Steps = c(adjSummaryByDate$TotalSteps, adjSummaryByDate$TotalStepsAdj))
ggplot(data = forGraph) + facet_grid(TypeData~.) +
  geom_histogram(aes(x = Steps), color = 'black', fill = 'light green')
```

Following code will print mean, median, standard deviation of steps in both scenario
```{r}
readingAdj[,list(Mean_Without_Adj = mean(steps, na.rm=TRUE), Mean_With_Adj = mean(AdjSteps),
                 Median_Without_Adj = median(steps, na.rm=TRUE), Median_With_Adj = median(AdjSteps),
                 Std_Dev_Without_Adj = sd(steps, na.rm=TRUE), Std_Dev_With_Adj = sd(AdjSteps))]
```
From above table and histogram, we can see that imputing missing value changes behavior of the data.

Total number of steps taken in earlier case were **`r sum(readingAdj$steps, na.rm=TRUE)`**  
Total number of steps taken in after adjustment were **`r sum(readingAdj$AdjSteps, na.rm=TRUE)`**
- - -

### Part 4: Are there differences in activity patterns between weekdays and weekends?
#### Question 1. Create a new factor variable in the dataset with two levels indicating whether a given date is a weekday or weekend day.
Following code will create factor from date
```{r}
weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
#Finding weekday or weekend
reading$Day = ifelse(weekdays(reading$date) %in% weekday, 'weekday', 'weekend')
rm(weekday)

reading$Day = as.factor(reading$Day)
```

#### Question 2. Make a panel plot containing a time series plot of the 5-minute interval against the average number of steps taken, averaged across all weekday days or weekend days.
Following code will generate average activity across weekend and weekday seperatly and then plot it.
```{r}
summaryByTimeByDay = reading[,list(Mean = as.numeric(mean(steps, na.rm = TRUE))),
                        by = list(Day,time)]

#Converting to POSIXct Time for ggplot use
summaryByTimeByDay$time = as.POSIXct(summaryByTimeByDay$time)

ggplot(summaryByTimeByDay) + geom_line(aes(x=time,y=Mean, color = Day), size = 1) + facet_grid(Day~.) +
  labs(x = 'Time', y = 'Average Number of Steps', title = 'Average daily activity pattern') +
  scale_x_datetime(labels = date_format("%H:%M")) + theme(legend.position="none")
```